Longitudinal Principal Manifold Estimation

### 2023-07-04 paper/main.tex
add data augmentation figure

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{geometry}
\geometry{
  left=10mm,
  right=10mm
  }

@@

To assess the performance of the LPME algorithm, we compare its performance with the that of two alternative approaches on simulated datasets. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered. The embedding functions used to define each manifold are given in Table \ref{table:1}. In these functions, the values of $\zeta$, $\alpha$, and $\beta$, as well as the functional form of $g(\cdot)$ are varied between trials. The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance varied between trials.

@@

In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}). In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$, when the spline smoothing method is used and $\lambda_g = g + 0.5$ for $g = 0, \dots, 25$ when the Gaussian process smoothing method is used. In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

Notably, the manifolds estimated in the simulated datasets included several connected manifolds, which methods based on the principal curves, including the principal manifold framework used here, are typically unable to fit due to the self-consistency property discussed above. To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, we the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically below [Figure still needs to be added].

>>>>

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}

@@

To assess the performance of the LPME algorithm, we compare its performance with the that of two alternative approaches on simulated datasets. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered. The embedding functions used to define each manifold are given in Table \ref{table:1}. In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials. The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance that is varied between trials. Meanwhile, in cases where $d=1$, $\alpha$ and $\beta$ are scalars that are set externally. If $d>1$, then $\alpha$ and $\beta$ take the form of vectors of length $d$, with the values drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.

@@

In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}). In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$. In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

Notably, the manifolds estimated in the simulated datasets included several connected manifolds, which methods based on the principal curves, including the principal manifold framework used here, are typically unable to fit due to the self-consistency property discussed above. To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, we the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:1}.

\begin{figure}%
  \centering
  \subfloat[\centering $D = 2$]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering $D = 3$]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Augmenting the Unit Circle}
  \label{fig:1}
\end{figure}
